---
title: "MGBC_SuppFig1"
author: "Benjamin Beresford-Jones"
date: "30/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required datasets:
- FINAL_bin_qc.tsv
- FINAL_bin_taxonomy.tsv
- s4_summary.tsv


## load libraries
```{r}
library(magrittr)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(reshape)
library(UpSetR)
library(tidyverse)
library(ggpol)
```

## set defaults
```{r}
INDIR="SuppFig1_data/" # unzip
# OUTDIR="CHANGEME" # change to save figures to specified directory

bS_cn <- c("Sample", "Species", "Reads_assigned", "Read_frac")
tax_cn <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")

f1_cn <- c("MGBC_id", "Orig_name", "Study", "Completeness", "Contamination", "Size", "N_contigs", "N50", "MCL", "QS", "CMseq", "Qual_level", "Lowest_tax_gtdb", "Lowest_tax_rank", "GTDB_taxonomy")

tax_order <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
```

## build colour palettes
```{r}
phycol=c("#377eb8","#8CB302","#008C74","#d95f02","#FF974F","#FFED6F","#FDCDAC","#ffd92f","#e22426","#B3B3B3","#FBB4AE","#984ea3","#35478C","#7FC97F","#FF73C5","#BF5693")
phylabs=c("Actinobacteriota","Bacteroidota","Campylobacterota","Cyanobacteria","Deferribacterota","Desulfobacterota","Elusimicrobiota","Firmicutes","Firmicutes_A","Firmicutes_B","Firmicutes_C","Proteobacteria","Spirochaetota","Thermotogota","Verrucomicrobiota","Verrucomicrobiota_A")
names(phycol) <- phylabs
```


## Load datasets
```{r}
# bin qc data
bin_qc <- read.delim(file = paste0(INDIR, "FINAL_bin_qc.tsv"), header = FALSE)

# bin_taxonomy
bin_taxonomy <- read.delim(file = paste0(INDIR, "FINAL_bin_taxonomy.tsv"), header = FALSE)
colnames(bin_taxonomy) <- c("Sample_id", "Binner", "Bin", "Taxonomy")

# core genome analysis data
cg <- read.delim(file = paste0(INDIR, "s4_summary.tsv"), header = FALSE, col.names = c("Species", "Binner", "Core_genome"))
```


## setup
```{r}
colnames(bin_qc) <- c("Sample_id", "Binner", "Bin", "Completeness", "Contamination", "GC", "Lineage", "N50", "Size", "Composite_binners")
bin_qc$Genome_id <- paste(bin_qc$Sample_id, bin_qc$Binner, bin_qc$Bin, sep = "_")
bin_qc$QS <- bin_qc$Completeness - (5*bin_qc$Contamination) # Calculate Quality Score 

# assign QC status to each genome
bin_qc$QC_level <- "FAIL"
bin_qc$QC_level[bin_qc$Completeness >= 50 & bin_qc$Contamination <= 5 & bin_qc$QS >= 50] <- "MQ" # medium-plus quality
bin_qc$QC_level[bin_qc$Completeness >= 90 & bin_qc$Contamination <= 5 & bin_qc$N50 >= 10000 & bin_qc$Size <= 8000000] <- "HQ"
```

## Analyses

# Number of bins - total, MQ and HQ
```{r}
# all genomes
all_qc <- table(bin_qc$Binner)
# MQ+ genomes
MQ_HQ_qc <- table(bin_qc$Binner[bin_qc$QC_level %in% c("MQ", "HQ")])
# HQ genomes
HQ_qc <- table(bin_qc$Binner[bin_qc$QC_level == "HQ"])

MQ_qc <- MQ_HQ_qc - HQ_qc
LQ_qc <- all_qc - MQ_HQ_qc

cat("Number of bins:\n")
cat("Of any quality:")
all_qc
cat("\nOf medium-plus quality:")
MQ_HQ_qc
cat("\nOf high quality:")
HQ_qc
```


```{r}
# Percentage of high quality bins of all bins
cat("Percentage of high quality bins of all bins:")
HQ_qc / all_qc * 100
# Percentage of high quality bins of all MQ bins
cat("\nPercentage of high quality bins of all medium-plus quality bins:")
HQ_qc / MQ_HQ_qc * 100
```

```{r}
mq_plus_qc <- bin_qc[bin_qc$QC_level %in% c("MQ", "HQ"),]
hq_plus_qc <- bin_qc[bin_qc$QC_level == "HQ",]
```


# Supplementary Figure 1a
```{r}
mq_plus_qc$Group <- "Medium quality"
hq_plus_qc$Group <- "High quality"
figure_qc <- hq_plus_qc

figure_qc$Purity <- 100-figure_qc$Contamination

colnames(figure_qc)[c(15,12)] <- c("Purity", "Quality Score")

figure_qc_m <- melt(figure_qc[,c("Binner", "Completeness", "QC_level", "Purity", "Quality Score")], id.vars = c("Binner", "QC_level"))

ggboxplot(figure_qc_m[figure_qc_m$QC_level == "HQ",], x = "Binner", y = "value", 
          fill = "variable", 
          palette = "Dark2",
          xlab = FALSE, ylab = "Value",
          outlier.shape = NA) +
  theme(legend.position = "right", legend.title = element_blank())
```

```{#r}
# save file
pdf(paste0(OUTDIR, "MGBC_S1-binner_quality_indices_comparison.pdf"), height = 3, width = 7)
ggboxplot(figure_qc_m[figure_qc_m$QC_level == "HQ",], x = "Binner", y = "value", 
          fill = "variable",
          palette = "Dark2",
          xlab = FALSE, ylab = "Value",
          outlier.shape = NA) +
  theme(legend.position = "right", legend.title = element_blank())
dev.off()
```

### Descriptive statistics

```{r}
hq_only <- figure_qc_m[figure_qc_m$QC_level == "HQ",]

# completeness
lapply(split(hq_only, hq_only$Binner), function(x) {
  c(summary(x$value[x$variable == "Completeness"]),
    Stdev.=sd(x$value[x$variable == "Completeness"]))
})
```

```{r}
# Purity
lapply(split(hq_only, hq_only$Binner), function(x) {
  c(100-summary(x$value[x$variable == "Purity"]),
    Stdev.=sd(x$value[x$variable == "Purity"]))
})
```

```{r}
# QS
lapply(split(hq_only, hq_only$Binner), function(x) {
  c(summary(x$value[x$variable == "Quality Score"]),
    Stdev.=sd(x$value[x$variable == "Quality Score"]))
})
```


# Supplementary Figure 1b
```{r}
bin_taxonomy$Genome_id <- paste(bin_taxonomy$Sample_id, bin_taxonomy$Binner, bin_taxonomy$Bin, sep = "_")

# merge taxonomic and qc profiles for bins
btm <- merge(x = bin_taxonomy, y = bin_qc, by = "Genome_id", all = FALSE)
btm <- btm[,-c(6:8)]
colnames(btm) <- gsub(pattern = "\\.x", replacement = "", colnames(btm))

# stratify taxonomy
tax_bd <- strsplit(btm$Taxonomy, split = ";") %>% do.call("rbind", .) %>% data.frame
colnames(tax_bd) <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
btm <- cbind(btm, tax_bd)
btm_HQ <- btm[btm$QC_level == "HQ",]

# get only known species
btm_species_HQ <- btm_HQ[grep(pattern = "s__.", btm_HQ$Species),]

# generate binner statistics for each binner
btm_s_HQ <- lapply(split(btm_species_HQ, f = btm_species_HQ$Species), function(x) {
  df <- lapply(split(x, x$Binner), function(y) {
    data.frame(Species=unique(y$Species),
               Binner=unique(y$Binner),
               Med=median(y$QS),
               Mean=mean(y$QS),
               Count=nrow(y))
  }) %>% do.call("rbind", .)
}) %>% do.call("rbind", .)
```

```{r}
min_bin <- lapply(split(btm_s_HQ, f = btm_s_HQ$Species), function(x) {
  x$Binner[which.min(x$Med)]
}) %>% do.call("rbind", .)

max_bin <- lapply(split(btm_s_HQ, f = btm_s_HQ$Species), function(x) {
  x$Binner[which.max(x$Med)]
}) %>% do.call("rbind", .)

bin_counts_HQ <- rbind(data.frame(BINNER=names(table(min_bin)), COUNT=as.numeric(table(min_bin)), MD="Ranked lowest binner"),
      data.frame(BINNER=names(table(max_bin)), COUNT=as.numeric(table(max_bin)), MD="Ranked highest binner"))
```

```{r}
bin_counts_HQ$COUNT[bin_counts_HQ$MD == "Ranked lowest binner"] <- bin_counts_HQ$COUNT[bin_counts_HQ$MD == "Ranked lowest binner"]*-1

bin_counts_HQ$MD <- factor(bin_counts_HQ$MD, levels = c("Ranked lowest binner", "Ranked highest binner"))

p <- ggbarplot(bin_counts_HQ, x = "BINNER", y = "COUNT", 
          fill = "MD", palette = "npg", ylab = FALSE) +
  facet_share(~MD, dir = "h", scales = "free", reverse_num = TRUE) + # note: scales = "free"
  coord_flip()
ggpar(p, ylab = "Number of species", legend = "none")
```

```{#r}
# save file
pdf(paste0(OUTDIR, "MGBC_S1b-best_worst_binner_species.pdf"), height = 2, width = 6)
ggpar(p, ylab = "Number of species", legend = "none")
dev.off()
```


# Supplementary Figure 1c

```{r}
list <- c()
for (i in 1:length(unique(btm_s_HQ$Binner))) {
  BINNER <- unique(btm_s_HQ$Binner)[i]
  list[[BINNER]] <- c(unique(btm_s_HQ$Species[btm_s_HQ$Binner == BINNER]))
}

p <- upset(fromList(list), order.by = "freq", mainbar.y.label = "Species Intersections", 
           sets.x.label = "Number of species", 
           sets.bar.color = get_palette(palette = "jco", 4)[c(4,3,1,2)])
p
```

```{#r}
# save file
pdf(paste0(OUTDIR, "MGBC_S1c-binner_upset.pdf"), height = 4, width = 6)
p
dev.off()
```


# Supplementary Figure 1d

```{r}
btm_species_HQ$Species <- gsub(pattern = "s__", replacement = "", btm_species_HQ$Species)

stest <- table(btm_species_HQ$Species) %>% sort(decreasing = TRUE) %>% head(n=12) %>% names

p <- ggboxplot(btm_species_HQ, x = "Species", y = "QS", 
               color = "black", fill = "Binner", palette = "jco",
               ylab = FALSE,
               select = stest, order = rev(stest), 
               outlier.shape = NA)

ggpar(p, rotate = TRUE, ylab = "Quality Score", legend = "none")
```

```{#r}
# save file
pdf(paste0(OUTDIR, "MGBC_S1d-species_quality_score.pdf"), height = 8, width = 6)
ggpar(p, rotate = TRUE, ylab = "Quality Score", legend = "none")
dev.off()
```


# Supplementary Figure 1e

```{r}
# isolate benchmark 
LM_icg=1403 # n=58
BG_icg=1210 # n=62
AM_icg=1791 # n=136
LJ_icg=1167 # 54

cg$Core_genome_perc <- 0
cg$Core_genome_perc[cg$Species == "LM"] <- cg$Core_genome[cg$Species == "LM"]/LM_icg*100
cg$Core_genome_perc[cg$Species == "BG"] <- cg$Core_genome[cg$Species == "BG"]/BG_icg*100
cg$Core_genome_perc[cg$Species == "LJ"] <- cg$Core_genome[cg$Species == "LJ"]/LJ_icg*100
cg$Core_genome_perc[cg$Species == "AM"] <- cg$Core_genome[cg$Species == "AM"]/AM_icg*100

cg$Species <- gsub(pattern = "LM", replacement = "Ligilactobacillus murinus", cg$Species) %>% 
  gsub(pattern = "BG", replacement = "Bifidobacterium globosum") %>%
  gsub(pattern = "LJ", replacement = "Lactobacillus johnsonii") %>%
  gsub(pattern = "AM", replacement = "Akkermansia muciniphila") 
```

```{r}
# absolute core genome size
ggbarplot(cg, x = "Species", y = "Core_genome", 
          position = position_dodge(0.8), add = "mean_ci",
               color = "black", fill = "Binner", palette = "jco", xlab = FALSE, ylab = "Core genome size")
```

```{r}
# percentage core genome size
p <- ggbarplot(cg, x = "Species", y = "Core_genome_perc", 
          position = position_dodge(0.8), add = "mean_ci",
               color = "black", fill = "Binner", palette = "jco", 
          xlab = FALSE, ylab = "% Isolate core genome") + 
  theme(legend.position = "right")

ggpar(p, x.text.angle = 60)
```

```{#r}
pdf(paste0(OUTDIR, "MGBC_S1e-species_core_genome_perc.pdf"), height = 5, width = 6)
ggpar(p, x.text.angle = 60)
dev.off()
```


## Descriptive statistics
### MetaBAT2 data
```{r}
lapply(split(cg, cg$Species), function(x) {
  summary(x$Core_genome_perc[x$Binner == "METABAT2"])
})
```

### MetaWRAP data
```{r}
lapply(split(cg, cg$Species), function(x) {
  summary(x$Core_genome_perc[x$Binner == "METAWRAP"])
})
```

