---
title: "MGBC_Fig5"
author: "Benjamin Beresford-Jones"
date: "29/05/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required datasets:
- all_drug_summary.tsv
- shared_rank_taxa.txt


## load libraries
```{r}
library(magrittr)
library(ggplot2)
library(ggpubr)
library(grid)
library(gridExtra)
```

## set defaults
```{r}
INDIR="Fig5_data/" # unzip
# OUTDIR="CHANGEME" # change to save figures to specified directory

bS_cn <- c("Sample", "Species", "Reads_assigned", "Read_frac")
tax_cn <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")

f1_cn <- c("MGBC_id", "Orig_name", "Study", "Completeness", "Contamination", "Size", "N_contigs", "N50", "MCL", "QS", "CMseq", "Qual_level", "Lowest_tax_gtdb", "Lowest_tax_rank", "GTDB_taxonomy")

dm_cn <- c("Reference", "Genome", "Pangenome_frac", "N_genomes", "Total_genomes", "Mean_seq_id", "Max_seq_id", "Min_seq_id", "Mean_bit_score", "Clus_90_rep", "Species", "Host", "Taxonomy")
```

## build colour palettes
```{r}
phycol=c("#377eb8","#8CB302","#008C74","#d95f02","#FF974F","#FFED6F","#FDCDAC","#ffd92f","#e22426","#B3B3B3","#FBB4AE","#984ea3","#35478C","#7FC97F","#FF73C5","#BF5693")
phylabs=c("Actinobacteriota","Bacteroidota","Campylobacterota","Cyanobacteria","Deferribacterota","Desulfobacterota","Elusimicrobiota","Firmicutes","Firmicutes_A","Firmicutes_B","Firmicutes_C","Proteobacteria","Spirochaetota","Thermotogota","Verrucomicrobiota","Verrucomicrobiota_A")
names(phycol) <- phylabs
```

## Load datasest
```{r}
dm <- read.delim(file = paste0(INDIR, "all_drug_summary.tsv"), header = F, col.names = dm_cn)
dm <- dm[!dm$Reference %in% c("Ddes_1357-CutC", "Ddes_1358-CutD"),] # remove non drug metabolising genes from analysese

shared_taxa <- read.delim(file = paste0(INDIR, "shared_rank_taxa.txt"), header=F,
                             col.names=c("Taxa")) %>% unlist %>% as.character
```

```{r}
tax_split <- strsplit(dm$Taxonomy, split = ";")

dm$Phylum <- lapply(tax_split, function(x){x[2]}) %>% unlist
dm$Class <- lapply(tax_split, function(x){x[3]}) %>% unlist
dm$Order <- lapply(tax_split, function(x){x[4]}) %>% unlist
dm$Family <- lapply(tax_split, function(x){x[5]}) %>% unlist
dm$Genus <- lapply(tax_split, function(x){x[6]}) %>% unlist

# 95% sequence id
dm_95 <- dm[dm$Mean_seq_id >= 95,]

dm_95_split <- split(dm_95, dm_95$Reference)
dm_split <- split(dm, dm$Reference)
```

# barplot of all species with seqid >= 95
```{r}
lapply(dm_95_split, function(x) {
  tmp_p <- lapply(split(x, f = x$Host), function(y) {
    y$Frac <- y$N_genomes/sum(y$N_genomes)*100
    tmp <- y
  }) %>% do.call("rbind", .)  
  
  tmp_p$Species[tmp_p$Frac < 5] <- "Other"
  
  ggbarplot(tmp_p, x = "Host", y = "Frac", fill = "Species", 
            title = unique(x$Reference)) +
    theme(legend.position = "right")
  
  })
```

# generate representative figures for paper
```{r}
# function to get above 5% species for colour palettes
get_bar_plot_species <- function(x) {
  tmp_p <- lapply(split(x, f = x$Host), function(y) {
  
    y$Gen_frac <- y$N_genomes/sum(y$N_genomes)*100
    
    y$Species[y$Gen_frac >= 5]

  }) %>% unlist %>% unique

  gsub(pattern = "s__", replacement = "", tmp_p)

}

# build palette
bp_species <- c(get_bar_plot_species(dm_95_split$BACDOR_03642),
                get_bar_plot_species(dm_95_split$BT_4096),
                get_bar_plot_species(dm_95_split$COLAER_00311))

bp_pal <- c(get_palette("Paired", k=length(bp_species)), "grey")
names(bp_pal) <- c(bp_species, "Other")

# function to generate bar plots
get_bar_plot <- function(x) {
  tmp_p <- lapply(split(x, f = x$Host), function(y) {
  
    y$Gen_frac <- y$N_genomes/sum(y$N_genomes)*100
    
    tmp <- y[y$Gen_frac >= 5,c(1,11,12,19)]
    
    tmp <- rbind(tmp,
                  data.frame(Reference=unique(y$Reference), Species="Other", Host=unique(y$Host), Gen_frac=100-sum(tmp$Gen_frac)))
    
  }) %>% do.call("rbind", .)  

  tmp_p$Species <- gsub(pattern = "s__", replacement = "", tmp_p$Species)

  ggbarplot(tmp_p, x = "Host", y = "Gen_frac", 
            fill = "Species", palette = bp_pal,
            title = unique(x$Reference)) +
    theme(legend.position = "right")

}

# run barplot function
get_bar_plot(dm_95_split$BACDOR_03642)
get_bar_plot(dm_95_split$BT_4096)
get_bar_plot(dm_95_split$COLAER_00311)
```

```{r}
# add an NA column for absent mouse species
add_rank <- dm_95_split$COLAER_00311[1,]
add_rank$Species <- NA
add_rank$Host <- "MOUSE"
pwl <- get_bar_plot(rbind(dm_95_split$COLAER_00311, add_rank))
pwl
```

```{#r}
# save files

# Figure 4b
pdf(paste0(OUTDIR, "MGBC_suppfig4_BACDOR_03642-snl.pdf"), height = 4, width = 4)
get_bar_plot(dm_95_split$BACDOR_03642)
dev.off()

# Figure 4a
pdf(paste0(OUTDIR, "MGBC_suppfig4_BT_4096-snl.pdf"), height = 4, width = 4)
get_bar_plot(dm_95_split$BT_4096)
dev.off()

# Figure 4c
pdf(paste0(OUTDIR, "MGBC_suppfig4_COLAER_00311-snl.pdf"), height = 4, width = 4)
pwl
dev.off()
```



# Descriptive analyses 

Q1: does the top human hit exist in the mouse:
```{r}
df <- lapply(dm_95_split, function(x) {
  h_tmp <- x[x$Host == "HUMAN",]
  m_tmp <- x[x$Host == "MOUSE",]
  
  H <- h_tmp[order(h_tmp$N_genomes, decreasing = TRUE)[1],]
  S <- ifelse(H$Species %in% m_tmp$Species, "Yes", "No")
  S2 <- ifelse(H$Species %in% shared_taxa, "Shared", "Not shared")
  
  data.frame(Reference=unique(x$Reference), 
             H_top_in_M=S, 
             Shared=S2)
}) %>% do.call("rbind", .)

table(df$H_top_in_M[df$Shared == "Shared"])
```

Where the most dominant species encoding a gene in the human host was a shared species with the mouse, the same gene product (≥95% sequence identity) was encoded by mouse-derived genomes for that species (27/27).

```{r}
df # show data
```

Q2: where the dominant human species is shared in mice, is this also the most dominant gene-encoding species in mice?

```{r}
df <- lapply(dm_95_split, function(x) {
  h_tmp <- x[x$Host == "HUMAN",]
  m_tmp <- x[x$Host == "MOUSE",]
  
  H <- h_tmp[order(h_tmp$N_genomes, decreasing = TRUE)[1],]
  M <- m_tmp[order(m_tmp$N_genomes, decreasing = TRUE)[1],]

  
  Ht <- c(H$Species, H$Genus, H$Family, H$Order, H$Class, H$Phylum)
  St <- ifelse(Ht %in% shared_taxa, "Shared", "Not shared")
  Mt <- c(M$Species, M$Genus, M$Family, M$Order, M$Class, M$Phylum)
  Tt <- c("Species", "Genus", "Family", "Order", "Class", "Phylum")
  
  Lt <- ifelse(Ht == Mt, "Yes", "No")
  
  data.frame(Feature=rep(unique(x$Reference), length(Ht)), 
             Human=Ht, 
             Mouse=Mt, 
             Comparison=Lt, 
             Rank=Tt,
             Shared_taxon=St)
}) %>% do.call("rbind", .)


df_s <- df[df$Rank == "Species" & df$Shared_taxon == "Shared",]
table(df_s$Comparison)
```

```{r}
df_s
```

However, in a third of the cases of shared species, the most dominant species that encoded these genes in the mouse host differed from the most dominant species in humans (9/27; 33%).

